---
title: Amuse-bouches from useR!2019
subtitle: Short examples from the R conference in Toulouse
date: '2019-07-18'
slug: amuse-bouches-from-user-2019
tags: 
  - conference
  - R
draft: false
url_source: https://github.com/sinarueeger/webpage/blob/master/blogdown/content/post/2019-07-16-amuse-bouche-from-user-2019.Rmd
url_code: https://github.com/sinarueeger/webpage/blob/master/blogdown/content/post/2019-07-16-amuse-bouche-from-user-2019.R
header:
  caption: ''
  image: ''
output:
  blogdown::html_page:
    toc: false
---

```{r 2019-07-16-amuse-bouche-from-user-2019-0, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
knitr::purl(input = "content/post/2019-07-16-amuse-bouche-from-user-2019.Rmd",
            output = "content/post/2019-07-16-amuse-bouche-from-user-2019.R")

```

```{r 2019-07-16-amuse-bouche-from-user-2019-1, message=FALSE, warning=FALSE, include=FALSE}
# library(showtext)
#suppressPackageStartupMessages(library(tidyverse))
knitr::opts_chunk$set(cache = TRUE)
```


The 3+ days at [useR!2019 in Toulouse](http://www.user2019.fr) were packed with great talks^[I cannot wait to see the recordings to catch up with the parallel sessions that I missed!] and good food - hence the [amuse-bouches](https://en.wikipedia.org/wiki/Amuse-bouche) word play.

Here are some R code bits from the conference. Hopefully convincing enough to start using a new package or change a workflow. Not everything was brand-new, but it was helpful to have someone talking through their inspiration and examples. 

Check out the speakers' materials - soon there will be recordings too. Some of the examples are also copied straight from the speakers' slide decks. 

1. [Tidy eval](#tidy-eval)
2. [usethis](#usethis)
3. [pak](#pak)
4. [Reshaping data](#reshaping-data)
5. [vroom](#vroom)
6. [data.table](#datatable)
7. [rray](#rray)

<!------------------ ------------------->
## 1. tidy eval
<!------------------ ------------------->

Speaker: [Lionel Henry](https://twitter.com/_lionelhenry) ([Slides](https://speakerdeck.com/lionelhenry/reusing-tidyverse-code))

I never warmed up to the bang-bangs and enquo's. Hence the new and more straight forward `{{ }}` (read: curly curly) for functional programming `{{ arg }}` in the tidyverse feel like a game-changer. 

For those more familiar with the previous framework: `{{ arg }}` is a shortcut for `!!enquo(arg)`.

### dplyr example

Let's say you have a dataset, here `iris`, and you want to compute the average `Petal.Length` for each `Species`:

```{r 2019-07-16-amuse-bouche-from-user-2019-2 }
library(dplyr)
iris %>% 
  group_by(Species) %>%
  summarise(avg = mean(Petal.Length, na.rm = TRUE))
```

You can use the "curly-curly" brackets if you want to turn this small bit of code into a function `group_mean()` with a `data`, `by` and `var` argument^[See [Slide 49+](https://speakerdeck.com/lionelhenry/reusing-tidyverse-code?slide=49).] (and want to pass the variables on in an unquoted way):

```{r 2019-07-16-amuse-bouche-from-user-2019-3 }
group_mean <- function(data, by, var) {
  
  data %>%
    group_by({{ by }}) %>%
    summarise(avg = mean({{ var }}, na.rm = TRUE))
}
```

We can then apply `group_mean()` to any dataset that has a grouping and a continuous variable, for example, the mammals sleep dataset in `ggplot2`:

```{r 2019-07-16-amuse-bouche-from-user-2019-4, echo=TRUE, results='hide'}
library(ggplot2)
group_mean(data = msleep, by = vore, var = sleep_total)

```


### ggplot2 example

Another common tidy eval application is `ggplot2`. In the example below, we want a customised plot: a scatterplot with a `geom_smooth` on top of it. 

```{r 2019-07-16-amuse-bouche-from-user-2019-5, message=FALSE, warning=FALSE}
library(ggplot2)
theme_set(theme_bw())
ggplot(data = iris, 
       aes(x = Sepal.Length, y = Petal.Length, group = Species, color = Species)
       ) +
  geom_point() +
  geom_smooth(method = "lm") + 
  ggtitle("Pack this plot into a function.")
```

Again, we can wrap the "curly-curly" brackets around the arguments and apply them to a different dataset. 

```{r 2019-07-16-amuse-bouche-from-user-2019-6, message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
plot_point_smooth <- function(data, x, y, gr = NULL, method = "lm") {
  
  ggplot(data = data, 
         aes({{ x }}, {{ y }}, group = {{ gr }}, color = {{ gr }})
         ) +
    geom_point() +
    geom_smooth(method = method)
  
}
```

```{r 2019-07-16-amuse-bouche-from-user-2019-6b, message=FALSE, warning=FALSE, echo=TRUE,fig.show = 'hide', results='hide'}
plot_point_smooth(msleep, x = sleep_total, y = sleep_rem, gr = NULL) + 
  ggtitle("Tidy eval with the msleep dataset")
```


<!------------------ ------------------->
## 2. usethis
<!------------------ ------------------->

Speaker: [Jenny Bryan](https://twitter.com/JennyBryan) ([Slides + Material + Demo](https://github.com/jennybc/2019-07_useR-toulouse-usethis))


```{r 2019-07-16-amuse-bouche-from-user-2019-7, message=FALSE, warning=FALSE}
# install.packages("usethis")
library(usethis)
```

Once upon a time, there was the package `devtools`. Then `devtools` became too large, and now the `usethis` package is taking over some of the convenience functions for workflows. 

`usethis` is all about avoiding to copy+pasting. For example, there is a function to edit the `.Rprofile` called `usethis::edit_r_profile()`. Whenever there is a slightly complicated task ahead (say _restarting R_), the `usethis` package will talk you through the whole process. 

There are lots of `use_*` to add or modify something to/in a project/package and three functions to create a package, a project or a github fork: 

1. `create_package()`
1. `create_project()`
1. `create_from_github()`


### Create a package

If you want to create a package, do the following (see also [screencast](https://www.youtube.com/watch?v=3vsPMyTfT8I&feature=youtu.be)): 

```{r 2019-07-16-amuse-bouche-from-user-2019-8, eval=FALSE, include=TRUE, echo=TRUE, results='hide'}
## 1. create the package skeleton
create_package("~/tmp/mypackage")

## 2. use git
use_git()

## 3. add a license
use_mit_license()

## 4. run check
# install.packages("devtools")
devtools::check()

## 5. commit all files with git

## 6. set up git + github
use_github()
## will update the DESCRIPTION file

## 7. install the package
devtools::install()

## 8. add a rmarkdown readme file
use_readme_rmd()
## knit + commit + push

## 9. clean up if this was only a demo
## install.packages("fs")
## fs::dir_delete("~/tmp/mypackage")
```


<!------------------ ------------------->
## 3. pak
<!------------------ ------------------->

Speaker: [Gábor Csárdi](https://twitter.com/GaborCsardi) ([Slides](https://github.com/gaborcsardi/pak-talk))

```{r 2019-07-16-amuse-bouche-from-user-2019-9 }
# install.packages("pak") ## or
# devtools::install_github("r-lib/pak")
```

It seems like `pak` will make package installation - conventional and for projects - more intuitive. Before installing anything, `pak` will give you a heads up on what will be installed or if there are any conflicts. 

`pak` has two main functions: `pak::pkg_*` and `pak:::proj_*`


### Conventional package installation

Play around with `usethis`^[See [demo 1](https://github.com/gaborcsardi/pak-talk/blob/master/pkg.Rmd)] and see what happens:

```{r 2019-07-16-amuse-bouche-from-user-2019-10, eval = FALSE, echo=TRUE, results='hide'}
pak::pkg_install("usethis")
pak::pkg_remove("usethis")
pak::pkg_install("r-lib/usethis")
pak::pkg_status("usethis")
```



### Package installation for projects

First, create a project with `usethis`, then install R packages directly into the project. 

```{r 2019-07-16-amuse-bouche-from-user-2019-11, eval=FALSE, include=TRUE, echo=TRUE, results='hide'}
usethis::create_project("~/tmp/test")

## check the directory
dir()

## initialise a dedicated R packages folder
pak:::proj_create()

## check the directory again
dir()

## check the DESCRIPTION file
readLines("DESCRIPTION")

## install usethis
pak:::proj_install("usethis")

## this installs dependencies into a private project library
readLines("DESCRIPTION")

## remove the project folder again
fs::dir_delete("~/tmp/test")
```



<!------------------ ------------------->
## 4. Reshaping data
<!------------------ ------------------->

Speaker: [Hadley Wickham](https://twitter.com/hadleywickham) ([Demo](https://gist.github.com/hadley/eb5c97bfbf257d133a7337b33d9f02d1))

```{r 2019-07-16-amuse-bouche-from-user-2019-12b , echo=TRUE, results='hide'}
# install.packages("tidyverse/tidyr")
library(tidyr)
```

What a history reshaping data in R already has! From `reshape` to `melt` + `cast`, over to `gather` + `spread` and now `pivot_long` + `pivot_wide`. Reshaping data stays a mind-bending task, but hopefully, these `pivot_*` functions will make life easier.

```{r 2019-07-16-amuse-bouche-from-user-2019-12}
# devtools::install_github("chrk623/dataAnim")
# Master's Thesis project by Charco Hui
library(dataAnim)

## Our two toy datasets
datoy_wide
datoy_long
```

Let's reshape the datasets^[See [demo](https://gist.github.com/hadley/eb5c97bfbf257d133a7337b33d9f02d1)]:

```{r 2019-07-16-amuse-bouche-from-user-2019-12c, echo=TRUE, results='hide'}

## lets make it longer
datoy_wide %>%
  pivot_longer(-Name, names_to = "Subject", values_to = "Score")

## lets make it wider
datoy_long %>%
  dplyr::mutate(Time = 1:nrow(datoy_long)) %>%
  pivot_wider(names_from = "Subject", values_from = c("Score", "Time"))

```


<!---------------- -------------------->
## 5. vroom
<!---------------- -------------------->

Speaker: [Jim Hester](https://twitter.com/jimhester_) ([Slides](https://speakerdeck.com/jimhester/vroom
) + [Screencast](screen cast: bit.ly/vroom-yt))

```{r 2019-07-16-amuse-bouche-from-user-2019-13, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
## install.packages("vroom")
library(vroom)
```

Importing large datasets into R can be a painful task. Especially if you only need a subset of the columns. And apparently, our thoughts drift off after 10 sec^[I can totally confirm that.] staring at the screen where it is _still_ loading the dataset.

`data.table::fread()` is always here to help. But now comes `vroom`!

### Get some large'ish data

First, we need some large dataset. To not burden our laptops too much^[If you can, choose the [UKBB + GIANT meta analysis results](https://portals.broadinstitute.org/collaboration/giant/index.php/GIANT_consortium_data_files#2018_GIANT_and_UK_BioBank_Meta-analysis), which are pretty large.], we will go for some exome based [GWAS](https://en.wikipedia.org/wiki/Genome-wide_association_study) results.

```{r 2019-07-16-amuse-bouche-from-user-2019-14}

## Source: https://portals.broadinstitute.org/collaboration/giant/index.php/GIANT_consortium_data_files

path_to_file_1 <- "Height_AA_add_SV.txt.gz"
path_to_file_2 <- "BMI_African_American.fmt.gzip" 


## Height
download.file(
  "https://portals.broadinstitute.org/collaboration/giant/images/8/80/Height_AA_add_SV.txt.gz",
  path_to_file_1)

## BMI
download.file(
  "https://portals.broadinstitute.org/collaboration/giant/images/3/33/BMI_African_American.fmt.gzip",
  path_to_file_2)


## File size
## install.packages("fs")
fs::file_size(path_to_file_1)
fs::file_size(path_to_file_2)
```

The two datasets have a mix of characters, numbers and decimals^[Apparently, characters are the most challenging ones for speed.].

### vroom vs DT

Here is how `vroom` works and a basic comparison to `data.table::fread` (I let you do the proper [benchmarking](http://adv-r.had.co.nz/Profiling.html) yourself).

```{r 2019-07-16-amuse-bouche-from-user-2019-15, message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
library(dplyr)

## With vroom
giant_vroom <- vroom::vroom(path_to_file_1)
giant_vroom_subset <- giant_vroom %>% select(CHR, POS) %>% filter(CHR == 1)

## The equivalent with data.table
giant_DT <- data.table::fread(path_to_file_1)
giant_DT_subset <- giant_DT %>% select(CHR, POS) %>% filter(CHR == 1)

```
    
### col_select for the win

```{r 2019-07-16-amuse-bouche-from-user-2019-16, message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
## Selecting columns
giant_vroom_select <- vroom::vroom(path_to_file_1, 
                                   col_select = list(SNPNAME, ends_with("_MAF")))
head(giant_vroom_select)

## Preventing columns from being imported
giant_vroom_remove <- vroom::vroom(path_to_file_1, 
                                   col_select = -ExAC_AFR_MAF)
head(giant_vroom_remove)

## Renaming on the fly
giant_vroom_rename <- vroom::vroom(path_to_file_1, 
                                   col_select = list(p = Pvalue, everything()))
head(giant_vroom_rename)
```

### Combining multiple datasets

```{r 2019-07-16-amuse-bouche-from-user-2019-17, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
data_combined <- vroom::vroom( 
                    c(path_to_file_1, path_to_file_2), 
                    id = "path")
table(data_combined$path)
```



<!------------------ ------------------->
## 6. data.table
<!------------------ ------------------->

Speaker: [Arun Srinivasan](https://twitter.com/arun_sriniv) ([Slides](https://t.co/TrvDVS07QD))

`data.table` has a pretty cool feature^[See [slides 26](https://t.co/TrvDVS07QD)]:

```{r 2019-07-16-amuse-bouche-from-user-2019-18 , echo=TRUE, results='hide'}
# install.packages("data.table")
library(data.table)

## Create a giant data.table
p <- 2e6
dat <- data.table(x = sample(1e5, p, TRUE), y = runif(p))

## Let's select a few rows
system.time(
  tmp <- dat[x %in% 2000:3000 ]
)

## do the same operation again
system.time(
  tmp <- dat[x %in% 2000:3000 ]
)

```




<!---------------- -------------------->
## 7. rray
<!---------------- -------------------->

Speaker: [Davis Vaughan](https://twitter.com/dvaughan32) ([Slides](https://github.com/DavisVaughan/2019-07-09_useR-2019-rray/blob/master/useR-2019-rray.pdf))

```{}
# devtools::install_github("r-lib/rray")
## may take some time
```

`rray` can do two things that are otherwise annoying/counter-intuitive in R:

- broadcasting (recycling dimensions)
- subsetting (`bag[,1, drop = FALSE]`)


### Matrices with base-r

Let's look at an example of matrix operations in base-r^[See also [slide 5](https://github.com/DavisVaughan/2019-07-09_useR-2019-rray/blob/master/useR-2019-rray.pdf).]. 


First, we want to add two matrices with similar dimensions:

```{r 2019-07-16-amuse-bouche-from-user-2019-19a, error=TRUE}

mat_1 <- matrix(c(15, 10, 8, 6, 12, 9), byrow = FALSE, nrow = 2) 
mat_2 <- matrix(c(5, 2, 3), nrow = 1)

## broadcasting won't work ❌
mat_1 + mat_2

```

Next, we want to select one matrix column:

```{r 2019-07-16-amuse-bouche-from-user-2019-19b, error=TRUE}
dim(mat_1[,2:3]) ## selecting two columns is fine

## subsetting won't preserve the matrix class ❌
dim(mat_1[,1]) ## why not 2x1?

length(mat_1[,1]) ## ah, it turned into a vector!

dim(mat_1[,1, drop = FALSE]) ## but with drop = FALSE we can keep it a matrix
```

### Matrices with rray

Let's do now the same task with `rray`.

```{r 2019-07-16-amuse-bouche-from-user-2019-20 , echo=TRUE, results='hide'}
library(rray)

(mat_1_rray <- rray(c(15, 10, 8, 6, 12, 9), dim = c(2, 3)))
(mat_2_rray <- rray(c(5, 2, 3), dim = c(1, 3)))

## Broadcasting works ✓
mat_1_rray + mat_2_rray

## Subsetting works ✓
dim(mat_1_rray[,2:3])
dim(mat_1_rray[,1])

## smart functions
mat_1_rray / rray_sum(mat_1_rray, axes = 1)
rray_bind(mat_1_rray, mat_2_rray, .axis = 1)
rray_bind(mat_1_rray, mat_2_rray, .axis = 2)

```





## More info

- Program + Slides: https://user2019.r-project.org/talk_schedule/
- Collection of slides during the conference by [Praer (Suthira Owlarn)](https://twitter.com/S_Owla): https://github.com/sowla/useR2019-materials
- Recordings by the [R Consortium on Youtube](https://www.youtube.com/channel/UC_R5smHVXRYGhZYDJsnXTwg) (keynotes available, rest soon to be published)

<!---------------- -------------------->
<!---------------- -------------------->
## Last but not least

The `rstatsmeme` package is a little gem discovered thanks to [Frie Preu](https://twitter.com/ameisen_strasse):

```{r 2019-07-16-amuse-bouche-from-user-2019-21, message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
# devtools::install_github("favstats/rstatsmemes")
library(rstatsmemes)
show_me_an_R_meme()
```

